A \texttt{shiny} New Opportunity for Interaction with Big Data in Undergraduate Education

### 2015-05-13 Body/ShinyDBSampler/ShinyDBSampler.Rnw
Edits- primarily grammatical to ShinyDBSampler chapter

<<config, echo=FALSE, eval=TRUE, include=FALSE>>=
# Preliminaries

@@

\begin{center}
\textbf{abstract}\\
As the availability of truly massive data sets proliferates it is enticing to incorporate these data sources into the curriculum of an undergraduate statistics course.  Major barriers exist for interacting with big data due to the computationally intense nature of working with large databases.  Difficulties include gaining access to the database, interacting with database management software and obtaining summary statistics or manageable subsamples from the database for student use.  This paper describes a web based application, the Shiny Database Sampler, which allows instructors to bypass these barriers using simple JavaScript based tools constructed using R and the R packages \texttt{shiny} and \texttt{RMySQL}. The Shiny Database Sampler allows instructors and/or students to obtain smaller subsamples from databases, using a variety of random sampling schemes. 
\end{center}

@@

Statistics education has been rapidly evolving in the past decade with respect to undergraduate course curriculum and assessment. Technology has played the role as a catalyst for many of these major changes.  An important change involves how data is accessed and analyzed in the classroom.   The GAISE report \citep{GAISEcollege} laid out six recommendations on how to improve the teaching of introductory statistics; two of which urge statistics instructors to ``Use technology for developing conceptual understanding and analyzing data'' and to ``Use real data''.   There are many software tools and online repositories for instructors to access real data for use in the statistics classroom; such as the Data and Story Library \citep{DASL} and its Australian counterpart, OzDASL \citep{OzDASL}, the Data Archives of the Journal of Statistics Education \citep{JSErepo}, CAUSE Web Repository \citep{CAUSErepo}  and Many Eyes \citep{ManyEyes}. These technological tools are wonderful for accessing many real data sets but the majority of the data sets currently available are quite small in scale.

Finzer et al. argue that in an introductory level statistics curriculum ``(w)hat seems to us to be missing are data sets-especially large and highly multivariate data sets-that are ripe for exploration and conjecture driven by the students' intrigue, puzzlement and desire for discovery'' \citep[p.1]{Finzer2007} . Large, real data sources are becoming increasingly available, but tend to be less easily accessible. A major contributor to this trend is the Freedom of Information Act which to ensures that non-classified data from the United States Government is publicly available \citep{FOIA}.  The online government resources at \url{www.data.gov/} \citep{DataGov}, \url{www.census.gov/} \citep{Census}, \url{www.nhtsa.gov/} \citep{NHTSA} and \url{www.cdc.gov/} \citep{CDC} are all locations of massive data stores. Governmental data sets contain rich information related to many socially relevant issues, making them prime candidates for engaging student interest.

The goal of connecting undergraduate statistics students with big data sources requires careful consideration to implement.  \citet{jacobs2009} speaks of the difficulties associated with using, "data whose size forces us to look beyond the tried-and-true methods that are prevalent at that time", including scaling computing tasks, avoiding costs of sub-optimal storage schemes and parallel processing.

New data technologies are needed in order to allow introductory statistics students to interact with big data sources, such as the governmental databases. This paper discusses the construction and functionality of the Shiny Database Sampler; a web-based application that allows students to pull random samples from large databases. Uses of the application within a lab assignment and course project for an introductory statistics course are detailed.  Lastly the responses from a technology user survey of the Shiny Database Sampler from 265 students of an introductory statistics course are analyzed and discussed.

% \hh{I think the GOLD book would be better than the reference to the Italian journal} In his paper on graphics for large data,  \citet[][p.~129]{Unwin1999} states that ``(t)he definition of large in relation to data is always changing. A data set that required substantial high performance computing one year becomes easily analyzable on a laptop a few years later''. What constitutes ``small data'' or ``big data'' is constantly being redefined in the field of statistics as computation allows us to collect, store and manipulate larger and larger data sets, but what is consistent is the desire to be able to analyze big data. 
% 
% \hh{Include a few data taxonomies here - there is an early one by Huber, \\
% then one by Wegman: Huge Data Sets and the Frontiers of Computational Feasibility
% Edward J. Wegman
% Journal of Computational and Graphical Statistics
% Vol. 4, No. 4 (Dec., 1995), pp. 281-295\\
% I am sure that there are more recent ones. Include them in a table to show how quickly things are changing wrt the size of data.\\
% The overarching goal is that we want to have interesting and relevant data, and that means that it can't be a toy data set.
% }

% \hh{general outline of the intro should be something like
% \begin{itemize}
% \item: Motivation of what do we do?
% \item What has been done? - that's where the part of the lit review comes in
% \item Why is what we are doing relevant, i.e. where are the holes in the lit review - that can be mixed with the previous item
% \item Outline of the structure of the rest of the paper
% \end{itemize}}

@@

Exposing students to large data is tricky because after a certain size, issues with transferring, storing and accessing data become too unwieldy to be feasible with a student's personal computer. Remote databases and database querying software puts its use outside the realm of most undergraduate statistics courses for both students and instructors. The Shiny Database Sampler tool was constructed to provide access to data in large databases by allowing users to obtain and download data subsets of a more manageable size.

Most introductory statistics courses are focused on small sample methodology, so the Shiny Database Sampler is designed to allow students to take appropriately sized subsets from the databases to practice methodology learned in class. Working with a small static subset from the large database would let the remainder of the database go to waste. Instead the tool allows students to specify a random sampling scheme that will pull the subset from database dynamically. This is done to emphasize the role of random sampling techniques in data collection, an important concept within an introductory statistics curriculum.

@@

The Shiny Database Sampler is a JavaScript based online application created using the \texttt{shiny} package \citep{shiny} in the statistical computing language \texttt{R} \citep{R}. The Shiny package uses \texttt{R} code files to generate the graphical user interface that interacts with an \texttt{R} session running on the server.  This was used in combination with the \texttt{RMySQL} package \citep{RMySQL} to allow the \texttt{R} session on the server machine to query the database at the user's request via buttons on the graphical user interface.

The interface was designed with a focus on the quality of the software as an educational tool. The field of software development defines six attributes contributing to software quality: Functionality, reliability, usability, efficiency, maintainability, efficiency and portability (\citealt{bevan1997quality}; \citealt{manuel2002quality}). The portability of the Shiny Database Sampler is excellent as it can be accessed online through any JavaScript enabled web browser. The reliability and maintainability of \texttt{shiny} applications is stems from the proper implementation of \texttt{R} and \texttt{shiny} on the web server. The key consideration with respect to efficiency was to optimize the database querying using proper indexing so that sample retrieval occurred almost instantaneously \citep{schwartz2012high}.

The graphical user interface contains two tabs, each broken into two panels. The layout is designed for the user to select actions in the side panel and view the results in the main panel. The functionality and usability of the "Sample and Summarize" and "Visualize" tabs are discussed in the following subsections.

\begin{figure}[hbtp]
\centering
\includegraphics[keepaspectratio=TRUE,width=.95\textwidth]{ShinyDBSampler/SampleAndSummarizeWithPanelLabels.png}
\caption{Shiny Database Sampler Layout for "Sample and Summarize" Tab} 
\label{fig:samplesummarizetab}
\end{figure}

\subsubsection{The ``Sample and Summarize'' tab}
%\paragraph{The ``Sample and Summarize'' tab} \hfill\newline

A screenshot of the interface for the \textit{Sample and Summarize} tab, the is shown in Figure~\ref{fig:samplesummarizetab} below.  A sidebar panel which contains all the sampling options and controls and a main panel which contains the data table and brief summary of the sampled data set.

The sidebar panel contains several fields and buttons for selecting and executing a sampling plan.  At the top of the sidebar is a drop-down menu to select the database table from which the user wants to take a random subset.  The current version of the tool allows users to access %workout data from an Iowa State fitness club called the RecMilers (\url{www.recservices.iastate.edu/fitness/recmilers}), 
the 2001-2009 Fatality Analysis Recording System accident data from the National Highway Traffic Safety Administration (\url{www.nhtsa.gov/FARS}) and the Public Use Micro Sample data from the 2000 United States Census (\url{www.census.gov/}).

After selecting the database, the user can choose between taking a simple or stratified random subsample of data from the database. If the user chooses simple random sampling then must specify a sample size; whereas if the user chooses stratified random sampling the strata variable and number of samples per stratum need to be specified. Lastly the user sets the seed for the selection algorithm. Setting the seed may seem to contradict the intention to draw random subsets, however it is included after careful consideration. For classroom settings, it is often desirable that students work with identical data sets for consistency of class discussion and grading, which setting the same seed and the same sampling specifications will accomplish. Obtaining a random sample from the databases is still possible with the additional -- but fairly trivial -- step of first randomly generating a starting seed.

Once the sampling setup is specified, the user may click the ``Get My Sample!" button and the randomly selected subsample of the database will be obtained and displayed in the main panel of the interface. Note that the interface only keeps track of the more recently selected subsample, which we will refer to as the \textit{active data}. Lastly, the side panel contains the button to download the selected subsample to a local drive on the user's computer.  The data will be downloaded as comma separated values (csv) file to the default download folder on the user's computer.

The main panel of the Shiny Database Sampler interface displays a data table and a basic summary of each variable in the active data. When logging into the webpage a default sample is taken and displayed until a sample of the users choosing is selected. The data table is searchable, sortable and expandable which makes it easy for the user to take a quick peek at the variable names and values that have been selected. The basic summary statistics for each variable are also displayed in the main panel below the data table; those familiar with \textit{R} programming will quickly recognize this as the verbatim output of the \textit{summary} function in \textit{R}. In the case that stratified sampling was used to draw the data, the summary for each variable is broken down by strata. The displays in the main panel of the Shiny Database Sampler are not intended to be the location for any extensive analysis of the active data but instead a quick check that the data that were sampled are what the user intended to select.

\subsubsection{The ``Visualize'' tab}
%\paragraph{The ``Visualize'' tab} \hfill\newline

The \textit{Visualize} tab of the interface is designed to construct basic plots of the active data drawn in the \textit{Sample and Summarize} tab. Figure~\ref{fig:visualizetab} shows a screenshot of the layout of the \textit{Visualize} tab. The sidebar panel contains fields for specifying variables and variable types that will be plotted. Univariate plots can be created by selecting the response variable from a drop-down menu containing a list of all variables and the variable type. Plots may also display bivariate relationships by additionally selecting an explanatory variable and it's variable type. The variable can be specified as numerical or categorical, causing the default plotting to generate a histogram or barchart, respectively. Variable types are not automatically specified for variables in the database for the deliberate purpose of forcing student users to consider appropriate ways to display the data.

Once the plotting options are selected in the sidebar panel, the user may click the ``Make my Plot!" button to generate and display the plot in the main panel. Table~\ref{tab:defaultplottypes} shows the possible plot types that are created based on the options selected. Since the visualization are intended only for preliminary exploration, the plots have a default construction and labeling that are not able to be customized.

\begin{figure}[htbp]
\begin{center}
\includegraphics[keepaspectratio=TRUE,width=.99\textwidth]{ShinyDBSampler/Visualize.png}
\end{center}
\caption{Shiny Database Sampler Layout for "Visualize" Tab} 
\label{fig:visualizetab}
\end{figure}

@@

The Shiny Database Sampler was developed for applications in statistics education and was implemented in the initial stages of development within the curriculum of an introductory statistics course. The alpha version of the software was used within one section of Statistics 104: \textit{Introduction to Statistics} for a group lab assignment and a course project. Later, a beta version was extended to be used in the lab assignment for more sections of the same course. These two applications demonstrate possible uses of the Shiny Database Sampler.

@@

The lab that utilized the Shiny Database Sampler was designed for students to think critically about sampling approaches, then use the tool allowed to treat the large database as 
a population from which to obtain survey data. Students were asked to consider the following pair of hypothetical situations:

@@

\item Suppose now that our goal has changed.  Now we wish to investigate the association between age and state of residency. We want to compare the median ages for different states. We still have a budget that allows us to survey around 1040 people. To collect our sample we decide to take a stratified random sample of 20 residents from each state in the United States plus the District of Columbia and Puerto Rico. 
\end{enumerate}

In each scenario students were asked to discuss the choice of sampling scheme, and in particular to identify potential problems. The students used the Shiny Database Sampler tool to obtain a sample from the database containing a 1\% microsample of the 2010 U.S. Census, from which they estimated mean and median age of U.S. residents. This lab was written to ensure that sampling concepts were the primary focus, with the Shiny Database Sampler acting in a supporting role. In order to avoid (sporadic) clicking of buttons to obtain samples without ever stopping to consider why the sampling approach matters, we intentionally designed the assignment to invite students to carefully consider sampling options {\it before} using the tool. The entire lab assignment can be found in Appendix \ref{labappend}.

@@

A second application of the Shiny Database Sampler was as a data source for a capstone project for three sections of the Stat 104 class. Students were required to do work in small groups to complete a capstone project that took a statistical approach to answering a question of their choosing. Groups were allowed to pick their own data source in order to accommodate students' interests in the subject matter of project. The only specific requirement for the project was a written report that explained the collection of bivariate data and an analysis of the association -- included the appropriate plots, inference and interpretations to answer their question of interest.

Six groups chose to run a mock survey, using the databases available through the Shiny Database Sampler as a stand-in for a large population. These groups were required to demonstrate a strong argument for the sampling plan they used within the Shiny Database Sample. For instance, one group wanted to know if the proportion of fatal accidents involving a drunk driver was higher in California than in Iowa; as part of the report they needed to argue why taking a stratified random sample of 100 fatal accidents per state would produce better information to answer their question than a simple random sample. Groups that gathered random samples from the Census and Accidents databases knew that they had gathered information from real life sources and they seemed genuinely engaged in the results of their projects.

%%-----------------------------------------------------------------------------------
\subsection{User Survey for Software Assessment}

As discussed above, the Shiny Database Sampler was designed for student use on course assignments. The quality of the software was initially assessed using a pluralistic walk through \citep{nielsen1994}, where the developer met iteratively with both statistical novices and experts to test the functionality of the early versions of the software. The software improvements that followed this inspection created the beta version of the Shiny Database Sampler that was ready for student use.

The second stage of assessment for the Shiny Database Sampler was a user survey to learn student opinions about their experience from using the software during the lab assignment described above. Specifically, we are interested in three aspects of the sampler: we want to know if students find the tool easy to operate, if they see the connection to sampling concepts and if they find the data engaging. Exploring the student responses on these three topics of interest helps us to assess the quality of the Shiny Database Sampler as an educational software.

The ease of use of the Shiny Database Sampler is an considered an important attribute from both an educational psychology and a software development viewpoint. Cognitive load theory postulates that human capacity to process information is limited and that learning is composed of \textit{extraneous load}, effort to overcome obstructions to new knowledge, and \textit{germane load}, effort used to form new schema and integrate ideas with existing knowledge. \citet{muller2008} explain that ``Finding ways to increase germane load and minimize extraneous load has been a central pursuit of researchers under this paradigm." Application of cognitive load theory is extended to the setting of software in \textit{human centered design} \citep{oviatt2006}. The goal is to make interfaces for educational technologies intuitive and easy to operate in order to minimize the extraneous load; thus allowing more mental resources to be devoted to developing and integrating new knowledge. The intuitive construction of the user interface is important component-based software development, where the usability of a tool is defined by its learnability, understandability and operability \citep{manuel2002quality}.

Active learning requires a high cognitive load and if prior knowledge is lacking, more scaffolding is necessary to support learning \citep{muller2008}. Since sampling concepts are typically new for students, it is important that students clearly recognize sampling concepts in the interface so that it can be used as part of an assignment that is scaffolded for active learning. If students are able to identify the role of the Shiny Database Sampler as a tool for learning about random sampling, it is more easily integrated into the learning process.

Finally, assessing student engagement with the data sources in the Shiny Database Sampler is important because higher student engagement is linked to higher academic performance and learning \citep{carini2006student}. The GAISE guidelines recommend that technological tools should be used to help teach statistical concepts and that the use of real data is important for student engagement, hence we focus on these topics \citep{GAISEcollege}. \citet{Neumann2013} found that students consider real data more interesting and engaging. The hope is that students using the Shiny Database Sampler will find the real nationally collected data sources engaging.

Student responses to the survey were collected in an anonymous survey following the group lab assignment -- as described in Section~\ref{LabOverview} -- that required students of Stat 104, Introduction to Statistics, at Iowa State University to use the Shiny Database Sampler tool. Six sections of Stat 104 students were surveyed. The students were informed that the survey was not required and that no penalties or rewards were affiliated with its completion.  Of the 320 students attending lab, 265 completed the survey.

@@

After completing the lab assignment, students were asked to fill out a survey consisting of twelve statements (referred to as {\it items} in the following, see Table \ref{tab:surveyquestions} for an overview). For each statement, students were asked for feedback on their level of agreement on a Likert scale from strongly disagree to strongly agree. 
The twelve items were designed to assess student opinion within three topics of four items each: ease of use, connection to sampling concepts, and engagement with the census data.  We will refer to these as the Ease, Concept and Engagement item sets.  For each group of four items, two were worded positively and two were worded negatively. Introducing negation with half of the items was done to reduce the response bias associated with \textit{acquiescence}, the tendency to respond positively irrespective of the item content due \citep{Furnham1986}. Responses were scored as -2 (strongly disagree), -1 (disagree), 0 (Neutral), 1 (agree), 2 (strongly agree).  Responses for negatively worded items were reverse-scored for the purposes of analysis,

@@

From Table \ref{tab:surveyquestions} we see that all response averages are positive after reverse-scoring. With the Ease items this indicates that students tend to find the tool relatively easy to operate. For frame of reference, we assume that students are comparing the difficulty of use with other educational technologies and webpages they have encountered in the past; in particular the \texttt{JMP} software used previously on their Stat 104 labs and homework. Students also tend to respond to Concept items in a manner that is affirmative that the tool connects them to sampling concepts. Students' responses are near to neutral for most items about engagement with the census data, with the exception of Engagement item 3.  The phrasing of this question seems to have led students to reconsider their engagement level and led to a consistently more positive attitude.

@@

A fluctuation diagram visually displays a contingency table of a pair of variables as  the area of blocks on the bivariate grid of all possible response values. A diagonal heavy fluctuation diagram indicates strong agreement or {\it internal consistency} between responses of the two items.  Figure \ref{fig:fluctuationDiagrams} contains fluctuation diagrams for all item pairs within topic sets.    
We notice that most pairs of responses fall heavily along the diagonal and are primarily in the upper right of each diagram. This indicates that most items within sets have strong agreement and that the response values are generally neutral to positive for all items after reverse-scoring.
For the item pairs in the Concept topic set we see that fluctuation diagrams have slightly larger off diagonal trends than items within the other two sets, which indicates a lower level of internal consistency for Concept items than in the other two topic sets.

>>>>

<<config, echo=FALSE, eval=TRUE, cache=TRUE, include=FALSE>>=
# Preliminaries

@@

\begin{center}
\textbf{Status}: In Preparation for Submission to \textit{Technology Innovations in Statistics Education} (TISE)\\
\end{center}

\begin{center}
\textbf{Author}\\
Karsten Maurer, Iowa State University\\
\end{center}

\begin{center}
\textbf{Abstract}\\
\end{center}
As the availability of truly massive data proliferates, it is enticing to incorporate these data sets into the curriculum of an undergraduate statistics course.  Major barriers exist for interacting with big data due to the computationally intense nature of working with large databases.  Difficulties include gaining access to databases, interacting with database management software, and obtaining summary statistics or manageable subsamples from the database for student use.  This paper describes a web-based software application, the \textit{Shiny Database Sampler}, which allows instructors and students to bypass these barriers using a simple point-and-click interface constructed through \texttt{R} and the \texttt{R} packages \texttt{shiny} and \texttt{RMySQL}. The Shiny Database Sampler allows instructors and students to obtain subsamples from databases, using a variety of random sampling schemes. Application and evaluation of the software indicate that students find the interface easy to use, well connected to course concepts, and engaging through access to real data.

%-----------------------------------------------------------

@@

Statistics education has been rapidly evolving in the past decade with respect to undergraduate course curriculum and assessment. Technology has played the role as a catalyst for many of these major changes. An important change involves how data is accessed and analyzed in the classroom. The GAISE report \citep{GAISEcollege} laid out six recommendations on how to improve the teaching of introductory statistics; two of which urge statistics instructors to ``Use technology for developing conceptual understanding and analyzing data'' and to ``Use real data''. There are many software tools and online repositories for instructors to access real data for use in the statistics classroom. These include the Data and Story Library \citep{DASL} and its Australian counterpart, OzDASL \citep{OzDASL}, the Data Archives of the Journal of Statistics Education \citep{JSErepo}, CAUSE Web Repository \citep{CAUSErepo}, and Many Eyes \citep{ManyEyes}. These repositories are wonderful for accessing many real data sets but the majority of the data sets currently available are quite small in scale.

\citet{Finzer2007} argue that in curricula for introductory level statistics ``(w)hat seems to us to be missing are data sets-especially large and highly multivariate data sets-that are ripe for exploration and conjecture driven by the students' intrigue, puzzlement and desire for discovery'' \citep[p.1]{Finzer2007}. Large, real data sources are becoming increasingly available, but tend to be less easily accessible. A major contributor to this trend is the Freedom of Information Act which to ensures that non-classified data from the United States government is publicly available \citep{FOIA}. The online government resources at \url{www.data.gov/} \citep{DataGov}, \url{www.census.gov/} \citep{Census}, \url{www.nhtsa.gov/} \citep{NHTSA} and \url{www.cdc.gov/} \citep{CDC} are all locations of massive data stores. Governmental data sets contain rich information related to many socially relevant issues, making them prime candidates for engaging student interest.

The goal of connecting undergraduate statistics students with big data sources requires careful consideration to implement. \citet{jacobs2009} speaks of the difficulties associated with using, ``data whose size forces us to look beyond the tried-and-true methods that are prevalent at that time'', including scaling computational tasks, avoiding sub-optimal storage schemes and parallel processing.

New data technologies are needed in order to allow introductory statistics students to interact with big data sources, such as the governmental databases. This paper discusses the construction and functionality of the Shiny Database Sampler; a web-based application that allows students to pull random samples from large databases. Then, details of the implementation within a lab assignment and course project for an introductory statistics course are discussed. Lastly, the survey responses from 265 introductory statistics students whom used the Shiny Database Sampler for the lab assignment are analyzed to evaluate the software.

@@

Exposing students to larger and larger data is tricky because it becomes increasingly unwieldy to transfer, store and access data on student's personal computers. Access through remote databases and database querying languages is outside the realm of comfort for both students and instructors in most undergraduate statistics courses. The Shiny Database Sampler is a simplified graphical user interface that is designed to allow students to take appropriately sized subsets from the databases to practice small sample methodology taught in most introductory statistics courses. Working with a small static subset from the large database would let the remainder of the database go to waste. Instead the tool allows students to specify a random sampling scheme that will pull the subset from database dynamically. This is done to emphasize the role of random sampling techniques in data collection, an important concept within an introductory statistics curriculum.

@@

The Shiny Database Sampler is a JavaScript based online application created using the \texttt{shiny} package \citep{shiny} in the statistical computing language \texttt{R} \citep{R}. The Shiny package uses \texttt{R} code files to generate the graphical user interface in a web browser that interacts with an \texttt{R} session running on the server.  It is used in combination with the \texttt{RMySQL} package \citep{RMySQL} to allow the \texttt{R} session on the server machine to query the database at the user's request via buttons in the graphical user interface.

The interface was designed with a focus on the quality of the software as an educational tool. The field of software development defines six attributes contributing to software quality: functionality, reliability, usability, efficiency, maintainability, efficiency and portability (\citealt{bevan1997quality}; \citealt{manuel2002quality}). The Shiny Database Sampler is highly portable, as it can be accessed online through any JavaScript enabled web browser. The reliability and maintainability of \texttt{shiny} applications depend upon proper implementation of \texttt{R} and \texttt{shiny} on the web server. The key consideration with respect to efficiency is to optimize the database querying using proper indexing so that sample retrieval occurs almost instantaneously \citep{schwartz2012high}.

The graphical user interface contains two tabs, each broken into two panels. The layout is designed for the user to select actions in the side panel and view the results in the main panel. The functionality and usability of the \textit{Sample and Summarize} and \textit{Visualize} tabs are discussed in the following subsections.

\subsubsection{The \textit{Sample and Summarize} tab}
%\paragraph{The ``Sample and Summarize'' tab} \hfill\newline

A screenshot of the interface for the \textit{Sample and Summarize} tab is shown in Figure~\ref{fig:samplesummarizetab} below. A sidebar panel which contains all the sampling options and controls, and a main panel which contains the data table and brief summary of the sampled data set.

\begin{figure}[hbtp]
\centering
\includegraphics[keepaspectratio=TRUE,width=.95\textwidth]{ShinyDBSampler/SampleAndSummarizeWithPanelLabels.png}
\caption{Shiny Database Sampler Layout for \textit{Sample and Summarize} Tab} 
\label{fig:samplesummarizetab}
\end{figure}

The sidebar panel contains several fields and buttons for selecting and executing a sampling plan.  At the top of the sidebar is a drop-down menu to select the database table from which the user may take a random subsample.  The current version of the tool allows users to access the 2001-2009 Fatality Analysis Recording System accident data from the National Highway Traffic Safety Administration (\url{www.nhtsa.gov/FARS}) and the Public Use Micro Sample data from the 2000 United States Census (\url{www.census.gov/}).

After selecting the database, the user can choose between taking a simple or stratified random subsample of data from the database. If the user chooses simple random sampling, then they must specify a sample size; whereas if the user chooses stratified random sampling the strata variable and number of observations per stratum need to be specified. Lastly the user sets the seed for the selection algorithm. Setting the seed may seem to contradict the intention to draw random subsets, however it was included after careful consideration. For classroom settings, it is often desirable for students to work with identical data sets for consistency of class discussion and grading. This can be accomplished by setting the same seed and the same sampling specifications. Obtaining a random sample from the databases is still possible with the additional -- but fairly trivial -- step of first randomly generating a starting seed.

Once the sampling setup is specified, the user may click the \textit{Get My Sample!} button and the randomly selected subsample of the database will be obtained and displayed in the main panel of the interface. Note that the interface only keeps track of the more recently selected subsample, which we will refer to as the \textit{active data}. Lastly, the side panel contains the button to download the active data to a local drive on the user's computer.  The data will be downloaded as comma separated values (csv) file to the default download folder on the user's computer.

The main panel of the Shiny Database Sampler interface displays a data table and a basic summary of each variable in the active data. When logging into the webpage a default sample is taken and displayed until a sample of the users choosing is selected. The data table is searchable, sortable and expandable which makes it easy for the user to take a quick peek at the variable names and values in the active data. The basic summary statistics for each variable are also displayed in the main panel below the data table; those familiar with \textit{R} programming will quickly recognize this as the verbatim output of the \textit{summary} function in \textit{R}. In the case that stratified sampling was used to draw the data, the summary for each variable is broken down by strata. The displays in the main panel of the Shiny Database Sampler are not intended to be the location for any extensive analysis of the active data but instead a quick check that it was what the user intended to select.

\subsubsection{The \textit{Visualize} tab}
%\paragraph{The ``Visualize'' tab} \hfill\newline

The \textit{Visualize} tab of the interface is designed to construct basic plots of the active data that was drawn in the \textit{Sample and Summarize} tab. Figure~\ref{fig:visualizetab} shows a screenshot of the layout of the \textit{Visualize} tab. The sidebar panel contains fields for specifying variables and variable types that will be plotted. Univariate plots can be created by selecting the response variable from a drop-down menu containing a list of all variables and the variable type. Plots may also display bivariate relationships by additionally selecting an explanatory variable and it's variable type.

The variable types must be manually specified as numerical or categorical. Variable types are not automatically specified for variables in the database for the deliberate purpose of forcing student users to consider appropriate ways to display the data. Table~\ref{tab:defaultplottypes} shows the possible plot types that can be created based on the options selected.

Once the plotting options are selected in the sidebar panel, the user may click the \textit{Make my Plot!} button to generate and display the plot in the main panel. Since the visualizations are intended only for preliminary exploration, the plots have a default construction and labeling that are not able to be customized.

\begin{figure}[htbp]
\begin{center}
\includegraphics[keepaspectratio=TRUE,width=.99\textwidth]{ShinyDBSampler/Visualize.png}
\end{center}
\caption{Shiny Database Sampler Layout for \textit{Visualize} Tab} 
\label{fig:visualizetab}
\end{figure}

@@

The Shiny Database Sampler was developed for applications in statistics education and was implemented in the initial stages of development within the curriculum of an introductory statistics course. The alpha version of the software was used within one section of Statistics 104: \textit{Introduction to Statistics} for a group lab assignment and a course project. Later, a beta version was extended to be used in a lab assignment for six more sections of the same course. These two applications demonstrate possible uses of the Shiny Database Sampler.

@@

The lab that utilized the Shiny Database Sampler was designed for students to think critically about sampling approaches, then the software allowed them to treat the large database as a population upon which to conduct their mock survey. Students were asked to consider the following pair of hypothetical situations:

@@

\item Suppose now that our goal has changed.  Now we wish to investigate the association between age and state of residency. We want to compare the median ages for different states. We still have a budget that allows us to survey of 1040 people. To collect our sample we decide to take a stratified random sample of 20 residents from each state in the United States plus the District of Columbia and Puerto Rico. 
\end{enumerate}

In each scenario students were asked to discuss the choice of sampling scheme and to identify potential problems or difficulties. The students used the Shiny Database Sampler tool to obtain a sample from the database containing a 1\% microsample of the 2010 U.S. Census, from which they estimated mean and median age of U.S. residents. This lab was written to ensure that sampling concepts were the primary focus, with the Shiny Database Sampler acting in a supporting role. In order to avoid (sporadic) clicking of buttons to obtain samples without ever stopping to consider why the sampling approach matters, we intentionally designed the assignment to invite students to carefully consider sampling options {\it before} using the tool. The complete lab assignment can be found in Appendix \ref{labappend}.

@@

A second application of the Shiny Database Sampler was as an optional data source for a capstone project for students of Stat 104. Students were required to work in small groups to complete a capstone project that took a statistical approach to answering a question of their choosing. In order to accommodate students' interests in the subject matter of the project, groups were allowed to pick their own data source. The only specific requirement for the project was a written report that explained the collection of bivariate data and an analysis of the association -- included the appropriate plots, inference and interpretations to answer their question of interest.

Six of the groups chose to run a mock survey, using the databases available through the Shiny Database Sampler as a stand-in for a large population. These groups were required to demonstrate a strong argument for the sampling plan they used within the Shiny Database Sample. For instance, if a group wanted to know if the proportion of fatal accidents involving a drunk driver was higher in California than in Iowa; as part of the report they needed to argue why taking a stratified random sample of 100 fatal accidents per state would produce better information to answer their question than a simple random sample. Groups that gathered random samples from the Census and Accidents databases knew that they had gathered information from real life sources and they seemed genuinely engaged in the results of their projects.

%%-----------------------------------------------------------------------------------
\subsection{User Survey for Software Evaluation}

As discussed above, the Shiny Database Sampler was designed for student use in course assignments. The quality of the software was initially assessed using a pluralistic walk through \citep{nielsen1994}, where the developer met iteratively with both statistical novices and experts to test the functionality of the early versions of the software. The software improvements that followed this inspection created the beta version of the Shiny Database Sampler that was ready for student use.

The second stage of evaluation for the Shiny Database Sampler was a user survey to learn student opinions about their experience from using the software during the lab assignment described above. Specifically, we are interested in three aspects of the software: we want to know if students find the tool easy to operate, if they see the connection to sampling concepts and if they find the data engaging. Exploring the student responses on these three topics of interest helps us to assess the quality of the Shiny Database Sampler as an educational software.

The ease of use of the Shiny Database Sampler is an considered an important attribute from both an educational psychology and a software development viewpoint. Cognitive load theory postulates that human capacity to process information is limited and that learning is composed of \textit{extraneous load}, effort to overcome obstructions to new knowledge, and \textit{germane load}, effort used to form new schema and integrate ideas with existing knowledge. \citet{muller2008} explain that ``Finding ways to increase germane load and minimize extraneous load has been a central pursuit of researchers under this paradigm." Application of cognitive load theory is extended to the setting of software in \textit{human centered design} \citep{oviatt2006}. The goal is to make interfaces for educational technologies intuitive and easy to operate in order to minimize the extraneous load; thus allowing more mental resources to be devoted to developing and integrating new knowledge. The intuitive construction of the user interface is an important component to software development, where the usability of a tool is defined by its learnability, understandability and operability \citep{manuel2002quality}.

Active learning requires a high cognitive load, and if prior knowledge is lacking more scaffolding is necessary to support learning \citep{muller2008}. Since sampling concepts are typically new for students, it is important that students clearly recognize sampling concepts in the interface; allowing it to be used as part of an assignment that is scaffolded for active learning. If students are able to identify the role of the Shiny Database Sampler as a tool for learning about random sampling, it is more easily integrated into the learning process.

Finally, assessing student engagement with the data sources in the Shiny Database Sampler is important because higher student engagement is linked to higher academic performance and learning \citep{carini2006student}. The GAISE guidelines recommend that technological tools should be used to help teach statistical concepts and that the use of real data is important for student engagement, hence we focus on these topics \citep{GAISEcollege}. \citet{Neumann2013} found that students consider real data more interesting and engaging. The hope is that students using the Shiny Database Sampler will find the real, nationally collected, governmental data sources engaging.

Student responses were collected in an anonymous survey following the group lab assignment -- as described in Section~\ref{LabOverview} -- that required students of an introductory statistics course, Stat 104, at Iowa State University to use the Shiny Database Sampler tool. Six sections of Stat 104 students were surveyed. The students were informed that the survey was not required and that no penalties or rewards were affiliated with its completion. Of the 320 students attending the lab, 265 completed the survey.

@@

After completing the lab assignment, students were asked to respond to a survey consisting of twelve statements (referred to as \textit{items} in the following, see Table~\ref{tab:surveyquestions} for an overview). For each statement, students were asked for feedback on their level of agreement on a Likert scale from strongly disagree to strongly agree. 
The twelve items were designed to assess student opinion within three topics of four items each: ease of use, connection to sampling concepts, and engagement with the census data.  We will refer to these as the \textit{Ease}, \textit{Concept} and \textit{Engagement} item sets.  For each group of four items, two were worded positively and two were worded negatively. Introducing negation with half of the items was done to reduce the response bias associated with \textit{acquiescence} -- the tendency to respond positively irrespective of the item content due \citep{Furnham1986}. Responses were scored as -2 (strongly disagree), -1 (disagree), 0 (neutral), 1 (agree), 2 (strongly agree).  Responses for negatively worded items were reverse-scored for the purposes of analysis.

@@

From Table~\ref{tab:surveyquestions} we see that all response averages are positive after reverse-scoring. With the Ease items this indicates that students tend to find the tool relatively easy to operate. For frame of reference, we assume that students are comparing the difficulty of use with other educational technologies and webpages they have encountered in the past; in particular the \texttt{JMP} software used previously on their Stat 104 labs and homework. Students also tend to respond to Concept items in a manner that is affirmative that the tool connects them to sampling concepts. Students' responses are near to neutral for most items about engagement with the census data, with the exception of Engagement item 3.  The phrasing of this question seems to have led students to reconsider their engagement level and led to a consistently more positive attitude.

@@

A fluctuation diagram is the visual analog of the contingency table, displaying frequency of each unique responses combination as the area of blocks on the bivariate grid containing all possible pairs of response values. A fluctuation diagram with large blocks on the diagonal indicates strong agreement or {\it internal consistency} between responses of the two items.  Figure~\ref{fig:fluctuationDiagrams} contains fluctuation diagrams for all item pairs within topic sets.    
We notice that most pairs of responses fall heavily along the diagonal and are primarily in the upper right of each diagram. This indicates that most items within sets have strong agreement and that the response values are generally neutral to positive for all items (after reverse-scoring the negative statements).
For the item pairs in the Concept topic set we see that fluctuation diagrams have slightly larger off-diagonal trends than items within the other two sets, which indicates a lower level of internal consistency for Concept items than in the other two topic sets.
